PAWpySeed: A parallelized Python/C package for analyzing DFT wavefunctions in the PAW formalism

### 2018-04-19 docs/paper.tex
trying a new overlapping spheres integration technique

\title{PAWpySeed: A parallelized Python/C package for analyzing
DFT wavefunctions in the PAW formalism}
\author{Kyle Bystrom\\
        kylebystrom@berkeley.edu\\
		UC Berkeley Department of Materials Science and Engineering}
\date{\today}
\maketitle
\abstract{A parallelized Python and C package has been developed
to perform analysis of density functional theory (DFT) wavefunctions in the
projector augmented wave (PAW) formalism. The primary function of this code
is to calculate the overlap operator of two single-particle Kohn-Sham states
from different periodic structures with the same lattice, which is made nontrivial
by the depency of PAW augmentation regions on the basis of the structure.
The code is potentially useful for the analysis of point defect levels in
solids because it can determine the composition of defect levels in the basis
of the single-particle states of the bulk system.}

@@

\subsection{Projector Augmented Wave (PAW) Method}

\subsection{Projections}
\begin{equation}
\ket{\psi_{n\mathbf{k}}}=\ket{\widetilde{\psi}_{n\mathbf{k}}}+
\sum_{a,n,l,m}(\ket{\phi_{a\ind}}-\ket{\widetilde{\phi}_{a\ind}})
\braket{\widetilde{p}_{a\ind}|\widetilde{\psi}_{n\mathbf{k}}}
\label{eq:1}
\end{equation}

\subsubsection{Projection onto Basis from Different Structure}
In the pure physical picture of a quantum mechanical system, the eigenfunctions of the time
independent Hamiltonian form a complete basis for any function in real space. By expressing
a wavefunction from a system in the basis of the eigenstates of another system, one might
be able to learn about the nature of the wavefunction in question. For example, projecting
the wavefunction of a defect level in a periodic crystal onto the eigenfunctions of the
bulk crystal might yield information about how the defect level is constructed. For example,
the defect state might project primarily onto the conduction band, characterizing it
as a weakly perturbed host state. Such an understanding of defect levels could help select
corrections to perform on the defect system.\cite{lany}
\begin{equation}
\centering
\begin{split}
\braket{\psi_{Rn_1\mathbf{k}}|\psi_{Sn_2\mathbf{k}}} & =\widetilde{O}+O_M+O_R+O_S+O_N\\
O_M & =\sum_{a\in M_{RS}}\sum_{l,m}\sum_{\epsilon_1}\sum_{\epsilon_2}
\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_{a\ind _1}}
(\braket{\phi_{a\ind _1}|{\phi_{a\ind _2}}}
-\bra{\widetilde{\phi}_{a\ind _1}}\ket{\widetilde{\phi}_{a\ind _2}})
\braket{\widetilde{p}_{a\ind _2}|\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_R & =\sum_{i\in N_R}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_i}
(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
\ket{\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_S & =\sum_{j\in N_S}\bra{\widetilde{\psi}_{Rn_1\mathbf{k}}}(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_N & =\sum_{i,j\in N_{RS}}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|
\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}
\end{split}
\label{eq:finp}
\end{equation}
Equation \ref{eq:finp} is derived in Appendix A.

@@

\begin{table}
\centering
\begin{tabular}{|c|c|c|}
\hline
Computational Task & $\Theta$ & Frequency*\\
\hline
$\widetilde{O}$ & $BKSW \sim n^2$ & per band\\
$O_M$ & $BKSNP \sim n^2$ & per band\\
$O_R$ and $O_S$ & $BKSNP \sim n^2$ & per band\\
$O_N$ & $BKSNP \sim n^2$ & per band\\
$\braket{\widetilde{p}_i|\widetilde{\psi}_{n\mathbf{k}}}$ & $BKSF\mathrm{log}(F)
\sim n^2\mathrm{log}(n)$ & per structure\\
$(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
\ket{\widetilde{\psi}_{n\mathbf{k}}}$ & $BKSNPW \sim n^3$ & per structure\\
spherical Bessel transform partial waves & $NPG\mathrm{log}(G) \sim n$ & per structure pair\\
project plane waves onto partial waves & $EPKW \sim n$ & per structure pair\\
projections for overlapping aug spheres & $NPL \sim n$ & per structure pair\\
\hline
\end{tabular}

\begin{tabular}{c|c}
Symbol & Definition\\
\hline
B & number of bands\\
E & number of elements\\
F & size of FFT grid\\
G & size of partial wave radial grid\\
K & number of k-points\\
L & size of Legendre-Gauss grid around ions\\
N & number of sites**\\
P & number of projector functions\\
S & number of spin states\\
W & number of plane waves\\
n & number of electrons (approximate scaling)
\end{tabular}
\caption{Runtime scaling functions for each component of the code and
definitions for shorthand symbols to express runtime. Approximate
scales with the number of electrons are also shown.
*The frequency refers to how often the routine is called. "Per band"
indicates that the routine runs once every time a band from one structure
is projected onto all the bands of a basis structure. "Per structure"
indicates that the routine is a "setup" routine that must be called once
for each structure and corresponding band structure to do perform projections.
"Per structure pair" is the same as "per structure" except that the routine
is only run once for a pair of structures with the same lattice and plane-wave
basis set, the bands of one such structure to be used as a basis set for the other.
**Number of sites flexibly refers to the number of sites relevant to the calculation,
which worst-case scales with the total number of sites in the structure. For example,
calculating $O_M$ and $O_N$ only require the sites in sets $M$ and $N_RS$, respectively.}
\label{tab:runtime}
\end{table}

@@

The all electron partial waves fluctuate rapidly near the nucleus,
so the pseudowavefunction cannot be projected onto the partial waves
via a uniform Cartesian grid as is doen with the projector functions.
However, A plane wave can be expanded as a sum of spherical
Bessel functions via Rayleigh expansion.
\begin{equation}
e^{i\mathbf{k} \cdot \mathbf{r}} = 4\pi \sum_{l=0}^{\infty}\sum_{m=-l}^{l}
i^l j_l(kr)Y_l^{m*}(\mathbf{\hat{k}})Y_l^m(\mathbf{\hat{r}})
\label{eq:pwexp}
\end{equation}
Therefore, the overlap operator between a partial wave and plane wave can be evaluated as a summation
of spherical Bessel function overlap integrals. The following definitions are
used in this derivation:

@@

>>>>

\title{PAWpySeed: A parallelized Python/C package for analyzing
DFT wavefunctions in the PAW formalism}
\author{Kyle Bystrom\\
        kylebystrom@berkeley.edu\\
		UC Berkeley Department of Materials Science and Engineering}
\date{\today}
\maketitle
\abstract{A parallelized Python and C package has been developed
to perform analysis of density functional theory (DFT) wavefunctions in the
projector augmented wave (PAW) formalism. The primary function of this code
is to calculate the overlap operator of two single-particle Kohn-Sham states
from different periodic structures with the same lattice, which is made nontrivial
by the dependency of PAW augmentation regions on the basis of the structure.
The code is potentially useful for the analysis of point defect levels in
solids because it can determine the composition of defect levels in the basis
of the single-particle states of the bulk system.}

@@

%\subsection{Projector Augmented Wave (PAW) Method}

\subsection{Projections}

The equation of a PAW wavefunction is\cite{blochl}
\begin{equation}
\ket{\psi_{n\mathbf{k}}}=\ket{\widetilde{\psi}_{n\mathbf{k}}}+
\sum_{a,l,m,\epsilon}(\ket{\phi_{a\ind}}-\ket{\widetilde{\phi}_{a\ind}})
\braket{\widetilde{p}_{a\ind}|\widetilde{\psi}_{n\mathbf{k}}}
\label{eq:1}
\end{equation}
where the sum is over the sites $a$ in the structure and a set of
angular and radial quantum nubers $l$, $m$, and $\epsilon$ at each site.
The conventional overlap operator used for these wavefunctions is
\begin{equation}
\widetilde{A}=A+\sum_i\sum_j\ket{\widetilde{p}_i}(\bra{\phi_i}A\ket{\phi_j}
-\bra{\widetilde{\phi_i}}A\ket{\widetilde{\phi_j}})\bra{\widetilde{p}_j}
\label{eq:loc}
\end{equation}
where $i$ and $j$ are abbreviations for a particular index set
$a,l,m,\epsilon$. This overlap operator requires that the augmentation
regions of the two wavefunctions are same. Since these augmentation regions
are defined by the sites in the structure, the structures must be the same,
so a new formalism is necessary to calculate overlap operators for wavefunctions
in different structures.

\subsubsection{Projection onto Basis from Different Structure}
In the pure physical picture of a quantum mechanical system, the eigenfunctions of the time
independent Hamiltonian form a complete basis for any function in real space. By expressing
a wavefunction from a system in the basis of the eigenstates of another system, one might
be able to learn about the nature of the wavefunction in question. For example, projecting
the wavefunction of a defect level in a periodic crystal onto the eigenfunctions of the
bulk crystal might yield information about how the defect level is constructed. For example,
the defect state might project primarily onto the conduction band, characterizing it
as a weakly perturbed host state. Such an understanding of defect levels could help select
corrections to perform on the defect system.\cite{lany}
For two structures (denoted R and S) with the same
lattice but different bases, the overlap operator of two Kohn-Sham single particle states
is described by the following terms:
\begin{equation}
\centering
\begin{split}
\braket{\psi_{Rn_1\mathbf{k}}|\psi_{Sn_2\mathbf{k}}} & =\widetilde{O}+O_M+O_R+O_S+O_N\\
O_M & =\sum_{a\in M_{RS}}\sum_{l,m}\sum_{\epsilon_1}\sum_{\epsilon_2}
\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_{a\ind _1}}
(\braket{\phi_{a\ind _1}|{\phi_{a\ind _2}}}
-\braket{\widetilde{\phi}_{a\ind _1}|\widetilde{\phi}_{a\ind _2}})
\braket{\widetilde{p}_{a\ind _2}|\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_R & =\sum_{i\in N_R}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_i}
(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
\ket{\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_S & =\sum_{j\in N_S}\bra{\widetilde{\psi}_{Rn_1\mathbf{k}}}(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_N & =\sum_{i,j\in N_{RS}}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|
\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}
\end{split}
\label{eq:finp}
\end{equation}
Here, $M_{RS}$ is the set of sites that are identical in structures R and S,
$N_R$ and $N_S$ are all other sites in these structures,
and $N_{RS}$ is teh set of all pairs of sites, one in structure R and one in structure
S, with overlapping augmentation spheres. Equation \ref{eq:finp} is derived in Appendix A.

@@

\begin{table}
\centering
\caption{Runtime scaling functions for each component of the code and
definitions for shorthand symbols to express runtime. Approximate
scales with the number of electrons are also shown.
*The frequency refers to how often the routine is called. "Per band"
indicates that the routine runs once every time a band from one structure
is projected onto all the bands of a basis structure. "Per structure"
indicates that the routine is a "setup" routine that must be called once
for each structure and corresponding band structure to do perform projections. "Per lattice"
operations are required once per unique lattice used.
"Per structure pair" is the same as "per structure" except that the routine
is only run once for a pair of structures with the same lattice and plane-wave
basis set, the bands of one such structure to be used as a basis set for the other.
**Number of sites flexibly refers to the number of sites relevant to the calculation,
which worst-case scales with the total number of sites in the structure. For example,
calculating $O_M$ and $O_N$ only require the sites in sets $M$ and $N_RS$, respectively.}
\label{tab:runtime}

\begin{tabular}{|c|c|c|}
\hline
Computational Task & $\Theta$ & Frequency*\\
\hline
$\widetilde{O}$ & $BKSW \sim n^2$ & per band\\
$O_M$ & $BKSNP \sim n^2$ & per band\\
$O_R$ and $O_S$ & $BKSNP \sim n^2$ & per band\\
$O_N$ & $BKSNP \sim n^2$ & per band\\
$\braket{\widetilde{p}_i|\widetilde{\psi}_{n\mathbf{k}}}$ & $BKSF\mathrm{log}(F)
\sim n^2\mathrm{log}(n)$ & per structure\\
$(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
\ket{\widetilde{\psi}_{n\mathbf{k}}}$ & $BKSNPW \sim n^3$ & per structure\\
spherical Bessel transform partial waves & $NPG\mathrm{log}(G) \sim n$ & per structure pair\\
project plane waves onto partial waves & $EPKW \sim n$ & per lattice\\
projections for overlapping aug spheres & $NPL \sim n$ & per structure pair\\
\hline
\end{tabular}

\begin{tabular}{c|c}
Symbol & Definition\\
\hline
B & number of bands\\
E & number of elements\\
F & size of FFT grid\\
G & size of partial wave radial grid\\
K & number of k-points\\
L & size of Legendre-Gauss grid around ions\\
N & number of sites**\\
P & number of projector functions\\
S & number of spin states\\
W & number of plane waves\\
n & number of electrons (approximate scaling)
\end{tabular}
\end{table}

@@

The all electron partial waves fluctuate rapidly near the nucleus,
so the pseudowavefunction cannot be projected onto the partial waves
via coarse Cartesian grid as is doen with the projector functions.
However, A plane wave can be expanded as a sum of spherical
Bessel functions via Rayleigh expansion.
\begin{equation}
e^{i\mathbf{k} \cdot \mathbf{r}} = 4\pi \sum_{l=0}^{\infty}\sum_{m=-l}^{l}
i^l j_l(kr)Y_l^{m*}(\mathbf{\hat{k}})Y_l^m(\mathbf{\hat{r}})
\label{eq:pwexp}
\end{equation}
Therefore, the overlap operator between a partial wave and plane wave can be evaluated as a summation
of spherical Bessel function overlap integrals. The following definitions are
used in this derivation:

@@

\section{Projector function evaluation}

The projector function overlaps with the plane-wave states are calculated using the same
algorithm as VASP.\cite{vasp,vasp2,vasp3,vasp4}

\section{Projections for Overlapping Augmentation Spheres}

For two different structures, it is possible for atoms to have augmentation spheres
that overlap but are not directly on top of each other, or to have atoms of different elements
with the same position. In these cases, the overlap integrals of the partial waves
must be evaluated numerically on a per structure pair basis. The current method is a relatively
crude discrete integral in 3D space, which likely harms the precision of the overlap integral evaluation.
More precise and efficient methods are being considered.

\section{Results and Example Usage}

To test the program for its primary purpose of defect level analysis, the projection
of a defect level wavefunction onto the valence band of the bulk is defined:
\begin{equation}
v_d=\sum_{\mathbf{k}}\sum_{i\in \mathbb{V}}|\braket{\psi_{Bn_i\mathbf{k}}|\psi_{Dn_d\mathbf{k}}}|^2
\label{eq:vproj}
\end{equation}
where $i$ are the band indices of valence bands $\mathbb{V}$ in the bulk structure, the $B$
subscript denotes the bulk band structure, and the $D$ subscript denotes the point defect
band structure. This value, as well as the similarly defined $c_d$ for the conduction band,
were calculate for several charge states of the Silicon single vacancy, Phosphorus substitutional,
and Boron substitutional. The results are plotted in Figure \ref{fig:plots}.
As expected, the phosphorous and boron substitutional defects show litting mixing
of valence and conduction character in the defect levels because they are very shallow
defects, whereas the deeper vacancy character is shown by the mixing of conduction
and valence character in three nearly degenerate bands in the band gap.

\section{Conclusions}

PAWpySeed provides useful utilities for reading and analyzing PAW wavefunctions
from VASP, particularly for point defect levels. However, the current capabilities
of PAWpySeed are primarily qualitative because of mediocre numerical precision
and a lack of quantitative applications of the code. One potential quantitative
application to be developed in the future is the perturbation theory correction
proposedby Lany and Zunger for correcting defect level energies.\cite{lany}
The main barriers to such a correction are currently the lack of adequate
description of the deep level character of the defects and the difference
between the single particle states of DFT and higher level-of-theory
single-particle states. Work to resolve such issues is currently in development.
Namely, code is being written to construct atomic vacuum states onto which
PAW wavefunctions can be projected. In addition, the SCDM-k decomposition method
will written into PAWpySeed to construct localized orbitals for energy
corrections.\cite{scdm,scdmk}
